---
title: "OxyMetaG_Graphs"
author: "Joey Lin"
date: "2026-06-05"
output: html_document
---
Today, I will try to make some graphs for the oxymetag data i have on the bumblebees. My first graph will be analyzing the oxygen level of bumblebee in Laguna Beach from March to August. I hypothesize that the level of oxygen will increase as the lifecycle of the bumblebee.
```{r}
library(tidyverse)
merged <- read_csv("Data/merged_oxymetag.csv")
```
```{r}
laguna <- merged %>%
  filter(year == 2022) %>%
  mutate(date_full = as.Date(paste0("2022", date), format = "%Y%m%d"))

ggplot(laguna, aes(x = date_full, y = Oxygen_level)) +
  geom_point(size = 3) +
  geom_smooth(method = "loess", se = TRUE) +
  labs(
    title = "Gut Microbiome Oxygen Level Over Time",
    subtitle = "Bombus samples from Laguna Beach Botanical Garden (2022)",
    x = "Date",
    y = "Predicted Oxygen Level (%)"
  ) +
  theme_minimal()
```
Running Regression to check significance.
```{r}
# Convert date to numeric for regression
laguna <- laguna %>%
  mutate(date_numeric = as.numeric(date_full))

# Run linear regression
model <- lm(Oxygen_level ~ date_numeric, data = laguna)
summary(model)

# Add log transformed date
laguna <- laguna %>%
  mutate(date_log = log(date_numeric))

# Run log regression
model_log <- lm(Oxygen_level ~ date_log, data = laguna)
summary(model_log)
```

Results: not signifciant

# Elevation and Oxygen Level
Now, I will do regression of elevation oxygen and elevation aerobe pfams
```{r}
ggplot(merged, aes(x = elev_masl, y = Oxygen_level)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Elevation vs Oxygen Level",
    x = "Elevation (masl)",
    y = "Oxygen Level"
  ) +
  theme_minimal()

# Linear regression
model_linear <- lm(Oxygen_level ~ elev_masl, data = merged)
summary(model_linear)

# Log regression
model_log <- lm(aerobe_pfams ~ log(Oxygen_level), data = merged)
summary(model_log)
```
Result: elevation does not seem to explain the variation of oxygen levels in bumblebees (low p value but low r^2 value).

#Try multiple regression to increase r^2
```{r}
model_multi <- lm(Oxygen_level ~ elev_masl + species_id + country, data = merged)
summary(model_multi)
```
Plot species vs Oxygen level
```{r}
ggplot(merged, aes(x = reorder(species_id, Oxygen_level, median, na.rm = TRUE), 
                   y = Oxygen_level, fill = country)) +
  geom_boxplot() +
  labs(
    title = "Gut Microbiome Oxygen Level by Species",
    x = "Species",
    y = "Predicted Oxygen Level (%)",
    fill = "Country"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Results: Species type seems to explain more variation (adjusted r2 = 0.4485). When we isolate elevation from other factors, it does not seem statistically signifant (r^2 = 0.99).
