---
title: "Summer_2"
author: "Joey Lin"
date: "2026-07-19"
output: html_document
---
# Goals:
1. Elevation vs Oxygen level by species with sample size of at least 3 per species (follow up from summer    1 plot with n>=5)
2. Boxplot of oxygen level vs species/subgenus (include all points in box) using smplot
3. Explore if pham ratios are consistent across species
4. Boxplot of queens vs workers/caste (male/female) *intraspecies (facet by species)*

Load the data
```{r}
library(tidyverse)
library(here)
metadata <- read_csv("../Data/merged_oxymetag.csv")
```

## 1. Elevation vs Oxygen level by species (n>=3)
*assign color palette for species*
```{r}
species_palette <- c(
  "rubicundus"            = "#AD2632",
  "funebris"              = "#D36800",
  "hortulanus"            = "#BC8700",
  "robustus"              = "#DEBA00",
  "rufocinctus"           = "#E6625F",
  "morrisoni"             = "#FF855D",
  "griseocollis"          = "#FEA124",
  "vandykei"              = "#5951BE",
  "flavifrons_dimidiatus"  = "#6D37A0",
  "flavifrons_flavifrons"  = "#4D218C",
  "bifarius"              = "#B54C9B",
  "bifarius_nearcticus"   = "#DA65AA",
  "sylvicola"             = "#9A54BA",
  "huntii"                = "#862A85",
  "mixtus"                = "#CD4E77",
  "vosnesenskii"          = "#7C64D4",
  "fervidus"              = "#6E9887",
  "pensylvanicus"         = "#00A59E",
  "sonorus"               = "#006674",
  "atratus"               = "#007FB1",
  "insularis"             = "#67C864",
  "nevadensis"            = "#A4A4A4"
)
```
*generate the graph*
```{r}
metadata %>%
  # Remove rows missing elevation or oxygen measurements
  filter(!is.na(elev_masl), !is.na(Oxygen_level)) %>%
  
  # Temporarily group by species to count them
  group_by(species_id) %>%       
  
  # EXCLUDE species from the graph if they have fewer than 10 samples
  filter(n() >= 3) %>%          
  
  # Ungroup so ggplot handles the data properly
  ungroup() %>%                  
  
  # Send this filtered subset straight to the graph
  ggplot(aes(x = elev_masl, y = Oxygen_level, color = species_id)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = species_palette) +
  labs(
    title = "Elevation vs. Oxygen Level by Bumblebee Species",
    subtitle = "Excluding species with a sample size of less than 5",
    x = "Elevation (meters above sea level)",
    y = "Oxygen Level (%)",
    color = "Species ID"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 10, color = "gray40", hjust = 0.5, margin = margin(b = 15)),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    legend.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  ) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed")
```
*Check if the slope is significantly negative*
```{r}
model_elev <- lm(Oxygen_level ~ elev_masl, data = metadata %>%
  filter(!is.na(elev_masl), !is.na(Oxygen_level)) %>%
  group_by(species_id) %>%
  filter(n() >= 3) %>%
  ungroup())

summary(model_elev)
```
The slope is statistically significantly negative at (alpha=0.05). The R^2 is small. Lets try to add speices as a control variable in the regression:
```{r}
model_species <- lm(Oxygen_level ~ elev_masl + species_id, data = metadata %>%
  filter(!is.na(elev_masl), !is.na(Oxygen_level)) %>%
  group_by(species_id) %>%
  filter(n() >= 3) %>%
  ungroup())

summary(model_species)
```
### Conclusion: Elevation is no longer significant after controlling for species. However, speices is highly significanty and accounts for way higher variation (R^2 = 0.529). One concern is that species can be highly correlated with elevation, causing my multiple regression test to be unrelaible (multicollinearity). 
### Concern: Maybe month can come into play here. Lets check:
```{r}
metadata <- metadata %>%
  mutate(month = as.integer(substr(date, 1, 2)))
model_month_species <- lm(Oxygen_level ~ month + species_id, data = metadata %>%
  filter(!is.na(Oxygen_level), !is.na(month), !is.na(species_id)))

summary(model_month_species)
```
Month is not associated with oxygen level after controlling for species. 

## 2. Boxplot of Oxygen level vs species/subgenus (include all data points)
*install smplot2*
install.packages("smplot2")
```{r}
library (smplot2)
metadata %>%
  filter(!is.na(Oxygen_level), !is.na(species_id)) %>%
  ggplot(aes(x = species_id, y = Oxygen_level, color = species_id)) +
  sm_boxplot() +
  labs(
    title = "Gut Microbiome Oxygen Level by Species",
    x = "Species",
    y = "Oxygen Level (%)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
```
*now do one with subgenus (color by country)*
```{r}
metadata %>%
  filter(!is.na(Oxygen_level)) %>%
  ggplot(aes(x = subgenus, y = Oxygen_level, color = country)) +
  sm_boxplot() +
  labs(
    title = "Gut Microbiome Oxygen Level by Subgenus",
    x = "Subgenus",
    y = "Oxygen Level (%)",
    color = "Country"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
```
Now, I will run a statistical test comparing subgenus
```{r}
model_subgenus <- aov(Oxygen_level ~ subgenus, data = metadata %>% 
                        filter(!is.na(Oxygen_level), !is.na(subgenus)))
summary(model_subgenus)

# Post-hoc test to see which subgenera differ
TukeyHSD(model_subgenus)
```
### Conclusion: Thoracobombus (columbian) have significantly lower oxygen levels than US subgenra. Cullumanobombus (columbian) also has statistically significant lower oxygen levels than Pyrobombus.
```{r}
library(car)
vif(model_species)
```
### Concern: I had a question about multicollinearity between species and elevation. Claude told me to run the Vif test. Conculsion: there is no pure mathematical collinearity problem. Maybe run PGLS just to make sure?


## 3. Pham ratio consistency
```{r}
metadata %>%
  filter(!is.na(ratio), !is.na(species_id)) %>%
  group_by(species_id) %>%
  filter(n() >= 3) %>%
  ungroup() %>%
  ggplot(aes(x = reorder(species_id, ratio, median, na.rm = TRUE), 
             y = ratio, color = species_id)) +
  sm_boxplot(outlier.shape = NA) +
  coord_cartesian(ylim = c(0, 100)) +
  labs(
    title = "Pfam Ratio by Bumblebee Species",
    subtitle = "Excluding species with fewer than 3 samples",
    x = "Species",
    y = "Aerobe:Anaerobe Pfam Ratio"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
```

### Observation: the plot is consistent with oxygen levels. Atratus and robustus have the lowest and consistent ratios. Huntii and vosnesenskii have the highest ratios and variability. 

```{r}
model_ratio_species <- aov(ratio ~ species_id, data = metadata %>%
  filter(!is.na(ratio), !is.na(species_id)) %>%
  group_by(species_id) %>%
  filter(n() >= 3) %>%
  ungroup())

summary(model_ratio_species)
TukeyHSD(model_ratio_species)
```
```{r}
model_ratio_subgenus <- aov(ratio ~ subgenus, data = metadata %>%
  filter(!is.na(ratio), !is.na(subgenus)))

summary(model_ratio_subgenus)
TukeyHSD(model_ratio_subgenus)
```
### Concern: The p-value is just above the 5% level (p=0.0642). The raw pfam ratio does not differ significantly by species. Also, the raw pfam ratio does not differ by subgenra (p=0.785). Maybe oxymetag uses not just the raw pfam ratio to calculate oxygen level. 

## 4. Boxplot of queens vs workers/caste (male/female) *intraspecies (facet by species)*

Identify species with both workers and queens: 
```{r}
metadata %>%
  count(caste)
metadata %>%
  filter(caste %in% c("queen", "worker")) %>%
  group_by(species_id, caste) %>%
  summarise(n = n()) %>%
  pivot_wider(names_from = caste, values_from = n) %>%
  filter(!is.na(queen) & !is.na(worker))
```
```{r}
metadata %>%
  filter(caste %in% c("queen", "worker"),
         species_id %in% c("flavifrons_flavifrons", "huntii")) %>%
  ggplot(aes(x = caste, y = Oxygen_level, color = caste)) +
  sm_boxplot() +
  facet_wrap(~species_id) +
  labs(
    title = "Oxygen Level by Caste within Species",
    x = "Caste",
    y = "Oxygen Level (%)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
```
### Observation: Flavifron queens have oxygen levels labeled as NAs. Only huntii produced visualizable results. Sample size is small, but queens have higher oxygen levels compared to workers.

T-test for differences:
```{r}
t.test(Oxygen_level ~ caste, data = metadata %>%
  filter(species_id == "huntii", caste %in% c("queen", "worker")))
```
### Conculsion: Queen bumblebees have more aerobic gut microbiomes than workers for the species Huntii (p=0.003)!
