---
title: "Summer_1"
author: "Joey Lin"
date: "2026-06-30"
output: html_document
---

Today, I will be incorporating the color palette Nickole gave me into my graphs. 
First, load the data.
```{r}
library(tidyverse)
library(here)
metadata <- read_csv("../Data/merged_oxymetag.csv")
```
Lets assign the colors to each species
```{r}
species_palette <- c(
  "rubicundus"            = "#AD2632",
  "funebris"              = "#D36800",
  "hortulanus"            = "#BC8700",
  "robustus"              = "#DEBA00",
  "rufocinctus"           = "#E6625F",
  "morrisoni"             = "#FF855D",
  "griseocollis"          = "#FEA124",
  "vandykei"              = "#5951BE",
  "flavifrons_dimidiatus"  = "#6D37A0",
  "flavifrons_flavifrons"  = "#4D218C",
  "bifarius"              = "#B54C9B",
  "bifarius_nearcticus"   = "#DA65AA",
  "sylvicola"             = "#9A54BA",
  "huntii"                = "#862A85",
  "mixtus"                = "#CD4E77",
  "vosnesenskii"          = "#7C64D4",
  "fervidus"              = "#6E9887",
  "pensylvanicus"         = "#00A59E",
  "sonorus"               = "#006674",
  "atratus"               = "#007FB1",
  "insularis"             = "#67C864",
  "nevadensis"            = "#A4A4A4"
)
```


The first graph I will make is to create an elevation and oxygen level graph with the new color palette. 
```{r}
# 4. Generate the Elevation vs Oxygen Level Scatter Plot
elevation_oxygen_plot <- metadata %>%
  # Filter out rows missing elevation or oxygen measurements to keep the graph clean
  filter(!is.na(elev_masl), !is.na(Oxygen_level)) %>%
  
  ggplot(aes(x = elev_masl, y = Oxygen_level, color = species_id)) +
  
  # Add points with slightly larger size and subtle transparency to handle overlapping points
  geom_point(size = 3, alpha = 0.8) +
  
  # Apply your custom manual color scale mapping
  scale_color_manual(values = species_palette) +
  
  # Set labels and titles
  labs(
    title = "Elevation vs. Oxygen Level by Bumblebee Species",
    x = "Elevation (meters above sea level)",
    y = "Oxygen Level (%)",
    color = "Species ID"
  ) +
  
  # Clean, modern minimalist theme
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, margin = margin(b = 15)),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    legend.title = element_text(face = "bold"),
    panel.grid.minor = element_blank() # Cleans up background grid line clutter
  )

# 5. Display the plot in your RStudio Plots panel
print(elevation_oxygen_plot)

# Optional: Uncomment the line below to save the graph directly to your computer as a high-res image
# ggsave("elevation_vs_oxygen_plot.png", plot = elevation_oxygen_plot, width = 10, height = 6, dpi = 300)
```
I want a graph with species that have at least 5 samples
```{r}
# 4. Filter and plot simultaneously without changing the original file
metadata %>%
  # Remove rows missing elevation or oxygen measurements
  filter(!is.na(elev_masl), !is.na(Oxygen_level)) %>%
  
  # Temporarily group by species to count them
  group_by(species_id) %>%       
  
  # EXCLUDE species from the graph if they have fewer than 10 samples
  filter(n() >= 5) %>%          
  
  # Ungroup so ggplot handles the data properly
  ungroup() %>%                  
  
  # Send this filtered subset straight to the graph
  ggplot(aes(x = elev_masl, y = Oxygen_level, color = species_id)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = species_palette) +
  labs(
    title = "Elevation vs. Oxygen Level by Bumblebee Species",
    subtitle = "Excluding species with a sample size of less than 5",
    x = "Elevation (meters above sea level)",
    y = "Oxygen Level (%)",
    color = "Species ID"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 10, color = "gray40", hjust = 0.5, margin = margin(b = 15)),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    legend.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```
Using the phylogeny, I will now split the species into three main groups with different shapes assigned. 
```{r}

# 3. Comprehensive color palette covering all 22 species
species_palette <- c(
  "rubicundus"            = "#AD2632",
  "funebris"              = "#D36800",
  "hortulanus"            = "#BC8700",
  "robustus"              = "#DEBA00",
  "rufocinctus"           = "#E6625F",
  "morrisoni"             = "#FF855D",
  "griseocollis"          = "#FEA124",
  "vandykei"              = "#5951BE",
  "flavifrons_dimidiatus"  = "#6D37A0",
  "flavifrons_flavifrons"  = "#4D218C",
  "bifarius"              = "#B54C9B",
  "bifarius_nearcticus"   = "#DA65AA",
  "sylvicola"             = "#9A54BA",
  "huntii"                = "#862A85",
  "mixtus"                = "#CD4E77",
  "vosnesenskii"          = "#7C64D4",
  "fervidus"              = "#6E9887",
  "pensylvanicus"         = "#00A59E",
  "sonorus"               = "#006674",
  "atratus"               = "#007FB1",
  "insularis"             = "#67C864",
  "nevadensis"            = "#A4A4A4"
)

# 4. Map shapes globally to categorical text groups for the legend layout
# 17 = Triangle, 19 = Circle, 15 = Square
shape_palette <- c(
  "morrisoni, griseocollis, nevadensis" = 17,
  "vosnesenskii & other species"         = 19,
  "fervidus, pensylvanicus, sonorus, atratus" = 15
)

# 5. Process data and plot on a single set of axes
metadata %>%
  # Clean out entries missing coordinates or oxygen readings
  filter(!is.na(elev_masl), !is.na(Oxygen_level)) %>%
  
  # Categorize each species into one of the 3 legend shape labels on the fly
  mutate(shape_group = case_when(
    species_id %in% c("morrisoni", "griseocollis", "nevadensis") ~ "morrisoni, griseocollis, nevadensis",
    species_id %in% c("fervidus", "pensylvanicus", "sonorus", "atratus") ~ "fervidus, pensylvanicus, sonorus, atratus",
    TRUE ~ "vosnesenskii & other species"
  )) %>%
  
  # Map color to species_id, but map shape to the broader text label group
  ggplot(aes(x = elev_masl, y = Oxygen_level, color = species_id, shape = shape_group)) +
  
  # Scatter points layer (alpha adds light transparency for overlapping points)
  geom_point(size = 3, alpha = 0.75) +
  
  # Explicitly map the color palette and custom shape assignments
  scale_color_manual(values = species_palette) +
  scale_shape_manual(values = shape_palette) +
  
  # Titles and layout labels
  labs(
    title = "Elevation vs. Oxygen Level across Bumblebee Species",
    x = "Elevation (meters above sea level)",
    y = "Oxygen Level (%)",
    shape = "Lineage Group"
  ) +
  
  # Modern theme styling
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, margin = margin(b = 15)),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 9),
    panel.grid.minor = element_blank()
  ) +
  
  # CRITICAL: Turn off the color legend entirely, keeping only the shape guide
  guides(color = "none")
```
